---
title: "Certifying an e-variable"
subtitle: "Reverse information projection, then branch and bound"
format:
  revealjs:
    embed-resources: true
    theme: simple
    css: search-slides.css
    html-math-method: mathml
    slide-number: false
    width: 1280
    height: 760
execute:
  echo: false
  warning: false
---

```{r setup}
#| include: false
library(ripr)
library(riprvis)

K <- 3L
N_TRIALS <- 20L
TOL <- 1e-9

family <- multinomial_family(n_trials = N_TRIALS, k = K)
```

```{r plurality-null}
#| include: false
# H0 = union_j {theta : theta_1 <= theta_j}: candidate 1 does not win outright.
# Each facet is a simplex with the tie point replacing e_1, and the two overlap
# where candidate 1 trails both rivals.
plurality <- null_model(
  family,
  lapply(2:K, function(j) {
    vertices <- diag(K)
    vertices[, 1L] <- replace(numeric(K), c(1L, j), 0.5)
    simplex_region(vertices = vertices)
  })
)

# A margin small enough that the null is not trivially far away: a wide one puts
# the projection in a corner and both the fit and the search end immediately.
q_plur <- matrix(c(0.40, 0.34, 0.26), ncol = 1L)
Q_plur <- family(as.vector(q_plur))
```

```{r plurality-fit}
#| include: false
#| cache: true
#| dependson: ["setup", "plurality-null"]
# `record_gap = TRUE` on both verbs: the sweep is not free, but it is what puts
# a gap and a maximiser on every step rather than every second one.
set.seed(1L)
state_plur <- ripr_init(
  Q_plur,
  plurality,
  control = ripr_control(snapshot = "all")
)
for (i in seq_len(30L)) {
  state_plur <- state_plur |>
    fw_step(record_gap = TRUE) |>
    em_step(record_gap = TRUE)
}
fit_plur <- ripr_finish(
  state_plur,
  reoptimise = TRUE,
  identify = TRUE,
  record_gap = TRUE
)
```

```{r plurality-certify}
#| include: false
#| cache: true
#| dependson: ["plurality-fit"]
X_plur <- likelihood(Q_plur, label = "Q") /
  likelihood(fit_plur$P_star, label = "P*")
nodes_plur <- certify_trace(X_plur, plurality, tol = TOL)
```

```{r medial-null}
#| include: false
# H0: no candidate reaches a majority. One cell, and it touches no vertex of
# the simplex, so the bisection has nowhere to run to.
medial <- null_model(
  family,
  list(simplex_region(
    vertices = cbind(c(0.5, 0.5, 0), c(0.5, 0, 0.5), c(0, 0.5, 0.5))
  ))
)

# One dirac per candidate leading, equally weighted, so Q is a genuine mixture
# rather than a single multinomial.
q_med <- cbind(c(0.6, 0.2, 0.2), c(0.2, 0.6, 0.2), c(0.2, 0.2, 0.6))
W_med <- finite_dist(components = q_med, weights = rep(1 / 3, 3L))
Q_med <- family(W_med)
```

```{r medial-fit}
#| include: false
#| cache: true
#| dependson: ["setup", "medial-null"]
# Alternating steps: Frank--Wolfe adds an atom, EM reweights what is there.
set.seed(2L)
state_med <- ripr_init(Q_med, medial, control = ripr_control(snapshot = "all"))
for (i in seq_len(30L)) {
  state_med <- state_med |>
    fw_step(record_gap = TRUE) |>
    em_step(record_gap = TRUE)
}
fit_med <- ripr_finish(
  state_med,
  reoptimise = TRUE,
  identify = TRUE,
  record_gap = TRUE
)
```

```{r medial-certify}
#| include: false
#| cache: true
#| dependson: ["medial-fit"]
X_med <- likelihood(Q_med, label = "Q") /
  likelihood(fit_med$P_star, label = "P*")
nodes_med <- certify_trace(X_med, medial, tol = TOL)
```

```{r cup-null}
#| include: false
# A Gaussian family: X ~ N(theta, I) with theta a point of the plane. The null
# is one unbounded polyhedron, a cup: four vertices along its floor and two
# recession rays running up its sides, so it is drawn in V-representation and
# recedes past the frame rather than ending at an edge.
gauss <- gaussian_family(dim = 2L)
cup <- null_model(
  gauss,
  list(polyhedron_region(
    vertices = cbind(c(-1, 0), c(-0.4, -0.6), c(0.4, -0.6), c(1, 0)),
    rays = cbind(c(-0.3, 1), c(0.3, 1))
  ))
)

# Five equally weighted atoms ringing the cup, all outside it: one under the
# floor, one off each rim, and one flanking each ray further up.
q_cup <- cbind(
  c(0, -1.6),
  c(-1.9, 0.3),
  c(1.9, 0.3),
  c(-2.3, 2.2),
  c(2.3, 2.2)
)
W_cup <- finite_dist(components = q_cup, weights = rep(1 / 5, 5L))
Q_cup <- gauss(W_cup)
```

```{r cup-fit}
#| include: false
#| cache: true
#| dependson: ["cup-null"]
# The sample space is continuous and the alternative is a mixture, so
# Gauss--Hermite is out and the expectation is Monte Carlo: draws from Q are
# frozen at init and every step, gap sweep and drawn field reuses them.
set.seed(3L)
state_cup <- ripr_init(
  Q_cup,
  cup,
  engine = mc_engine(1500L),
  control = ripr_control(snapshot = "all")
)
for (i in seq_len(20L)) {
  state_cup <- state_cup |>
    fw_step(record_gap = TRUE) |>
    em_step(record_gap = TRUE)
}
fit_cup <- ripr_finish(
  state_cup,
  reoptimise = TRUE,
  identify = TRUE,
  record_gap = TRUE
)
```

```{r payloads}
#| include: false
# Everything the widgets draw. The fits and certifications above are cached;
# this reshaping is cheap enough to run every render, and riprvis owns the
# R-to-browser crossing that the old ojs_define()/JSON plumbing did by hand.
lattice <- ripr_lattice_data(family)

plur_problem <- ripr_problem_simplex_data(
  plurality,
  q = q_plur,
  title = "plurality",
  part_labels = c("θ₁ ≤ θ₂", "θ₁ ≤ θ₃")
)
plur_fit <- ripr_fit_simplex_data(state_plur, lattice, q = q_plur)
plur_cert <- ripr_certify_simplex_data(nodes_plur, tol = TOL)

med_problem <- ripr_problem_simplex_data(
  medial,
  q = q_med,
  weights = rep(1 / 3, 3L),
  title = "medial",
  part_labels = "θⱼ ≤ ½ for every j"
)
med_fit <- ripr_fit_simplex_data(
  state_med,
  lattice,
  q = q_med,
  weights = rep(1 / 3, 3L)
)
med_cert <- ripr_certify_simplex_data(nodes_med, tol = TOL)

# The planar payloads: the cup's generators frame the view, and the field is
# evaluated in R on that frame's grid with the fit's own frozen draws.
cup_problem <- ripr_problem2d_data(
  cup,
  q = q_cup,
  weights = rep(1 / 5, 5L),
  title = "cup",
  part_labels = "conv(V) + cone(R)"
)
cup_fit <- ripr_fit2d_data(state_cup, cup_problem)
```

## The plurality problem {.slide}

::: {.stage}
::: {.left}
```{r}
ripr_problem_simplex(plur_problem, height = 470)
```
:::

::: {.right .statement}
Counts over three candidates,
$$Y \sim \mathrm{Multinomial}(n, \theta), \qquad \theta \in \Delta^2, \quad n = 20 .$$

The alternative is a single multinomial — one candidate leads by a small margin:
$$Q = \mathrm{Multinomial}(n, q), \qquad q = (0.40,\ 0.34,\ 0.26).$$

The null says candidate 1 does not win outright — a union of two facets,
overlapping where candidate 1 trails both rivals:
$$H_0 : P \in \mathcal{P}_0 = \bigcup_{j \neq 1}
   \{P_\theta : \theta_1 \leq \theta_j\} .$$

We want the reverse information projection of $Q$ onto the convex hull of
$\mathcal{P}_0$ — a hull over *measures*, since $\mathcal{P}_0$ itself is not
convex:
$$P^\star = \operatorname*{arg\,min}_{P \in \mathrm{conv}(\mathcal{P}_0)}
   D_{\mathrm{KL}}(Q \,\|\, P).$$
:::
:::

::: {.caption}
Grey is the null, the black point the support of $Q$. With $K = 3$ the parameter
is a point in the triangle, so the whole problem is drawable.
:::

## Fitting the projection {.slide}

```{r}
ripr_fit_simplex(plur_problem, plur_fit, lattice, height = 545)
```

::: {.caption}
$g_i(\theta) = E_\theta[Q/P_i]$, red where it exceeds one. Frank–Wolfe adds the
maximiser over the null as an atom; the gap is $\max_{H_0} G_i - 1$, so the fit
is done when no red is left inside the dashed region.
:::

## Certifying the fit {.slide}

```{r}
ripr_certify_simplex(plur_problem, plur_cert, height = 545)
```

::: {.caption}
Repeated bisection, one facet at a time — the two are searched independently.
The panel is that facet's enclosure $[\text{incumbent},\ \text{bound}]$
closing; the certified bound is the larger of the two facets'.
:::

## The medial-triangle problem {.slide}

::: {.stage}
::: {.left}
```{r}
ripr_problem_simplex(med_problem, height = 470)
```
:::

::: {.right .statement}
The same family,
$$Y \sim \mathrm{Multinomial}(n, \theta), \qquad n = 20 .$$

The alternative is now a *mixture* — each candidate leads in turn, equally
often:
$$Q = \int \mathrm{Multinomial}(n, \theta)\, \mathrm{d}W(\theta), \qquad
W = \tfrac{1}{3} \sum_{i=1}^{3} \delta_{q_i},$$
with $q_1 = (0.6,\ 0.2,\ 0.2)$ and $q_2, q_3$ its permutations.

The null says no candidate reaches a majority — the medial triangle, a single
region interior to the simplex:
$$H_0 : P \in \mathcal{P}_0 =
   \{P_\theta : \theta_j \leq \tfrac{1}{2} \ \text{ for every } j\} .$$

The projection is the same problem:
$$P^\star = \operatorname*{arg\,min}_{P \in \mathrm{conv}(\mathcal{P}_0)}
   D_{\mathrm{KL}}(Q \,\|\, P).$$
:::
:::

::: {.caption}
Three atoms of equal weight, each outside the null, and a null touching no
vertex of the simplex.
:::

## Fitting the projection {.slide}

```{r}
ripr_fit_simplex(med_problem, med_fit, lattice, height = 545)
```

::: {.caption}
Steps alternate: Frank–Wolfe adds the maximiser of $g_i$ over the null, then EM
reweights the atoms already there.
:::

## Certifying the fit {.slide}

```{r}
ripr_certify_simplex(med_problem, med_cert, height = 545)
```

::: {.caption}
One cell, interior to the simplex, so the bisection has no vertex to run to.
:::

## The cup problem {.slide}

::: {.stage}
::: {.left}
```{r}
ripr_problem2d(cup_problem, height = 470)
```
:::

::: {.right .statement}
A continuous family now — one Gaussian draw with known covariance,
$$X \sim \mathrm{N}(\theta, I_2), \qquad \theta \in \mathbb{R}^2 .$$

The alternative is a mixture of five, equally weighted, ringing the null:
$$Q = \int \mathrm{N}(\theta, I_2)\, \mathrm{d}W(\theta), \qquad
W = \tfrac{1}{5} \sum_{j=1}^{5} \delta_{q_j},$$
with $q_1 = (0, -1.6)$, $q_{2,3} = (\mp 1.9,\ 0.3)$ and
$q_{4,5} = (\mp 2.3,\ 2.2)$.

The null is a single unbounded polyhedron — a cup, four vertices along the
floor and two rays running up the sides:
$$H_0 : \theta \in \mathrm{conv}(V) + \mathrm{cone}(R), \qquad
V = \begin{pmatrix} -1 & -0.4 & 0.4 & 1 \\ 0 & -0.6 & -0.6 & 0 \end{pmatrix},
\quad
R = \begin{pmatrix} -0.3 & 0.3 \\ 1 & 1 \end{pmatrix} .$$

The projection is the same problem — $\mathcal{P}_0$ is convex in $\theta$,
but its hull over measures is still larger than $\mathcal{P}_0$ itself:
$$P^\star = \operatorname*{arg\,min}_{P \in \mathrm{conv}(\mathcal{P}_0)}
   D_{\mathrm{KL}}(Q \,\|\, P).$$
:::
:::

::: {.caption}
Grey is the null, receding past the frame along its rays; the black points
are the atoms of $Q$. The frame is sized to the vertices and the atoms, so
the geometry nearest the alternative fills the view however far the cup runs.
:::

## Fitting the projection {.slide}

```{r}
ripr_fit2d(cup_problem, cup_fit, height = 545)
```

::: {.caption}
The same alternation, with $g_i$ now a Monte Carlo expectation over draws
from $Q$ frozen for the whole fit — the drawn field uses the same draws, so
the gap markers agree with it by construction. There is no certifying slide:
the only bounding method is the Bernstein enclosure of a multinomial
expectation over a simplex, and this null is unbounded besides.
:::
