---
title: "Visualising a RIPr fit and its certification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualising a RIPr fit and its certification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_ripr <- requireNamespace("ripr", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_ripr
)
```

riprvis draws the three stages of building a certified e-variable with
[ripr](https://github.com/fleverest/ripr): the problem (a composite null and
a fixed alternative), the fit (the reverse information projection of the
alternative onto the null), and the certification (a branch-and-bound upper
bound on the worst-case null expectation). Each stage has two verbs: a
`vis_*()` htmlwidget for exploring, which works in the RStudio viewer, R
Markdown and Quarto documents, and Shiny; and a `plot_*()` static plot for
print.

This vignette needs `ripr` installed
(`remotes::install_github("fleverest/ripr")`); without it the code below is
shown but not run.

## The problem

The running example is the *plurality* null for a three-category multinomial
with 20 trials: candidate 1 does not win outright, a union of two overlapping
simplices. The alternative `Q` puts candidate 1 narrowly ahead.

```{r problem}
library(ripr)
library(riprvis)

family <- multinomial_family(n_trials = 20L, k = 3L)
plurality <- null_model(
  family,
  lapply(2:3, function(j) {
    vertices <- diag(3)
    vertices[1L, ] <- replace(numeric(3), c(1L, j), 0.5)
    simplex_region(vertices = vertices)
  })
)
q <- c(0.40, 0.34, 0.26)
labels <- c("θ₁ ≤ θ₂", "θ₁ ≤ θ₃")

vis_problem(plurality, alternative = family(q), part_labels = labels)
```

## Fitting the projection

Alternating Frank–Wolfe and EM steps, with snapshots of every iterate and the
duality gap swept at every step so the widget can show where the gap is
attained -- the starting mixture is a frame too, with its gap swept at
init.

```{r fit}
set.seed(1L)
state <- ripr_init(
  family(q), plurality,
  record_gap = TRUE,
  control = ripr_control(snapshot = "all")
)
for (i in seq_len(15L)) {
  state <- state |>
    fw_step(record_gap = TRUE) |>
    em_step(record_gap = TRUE)
}

vis_fit(state, part_labels = labels)
```

Drag the step slider (or press play) to watch the field
\(G_i(\theta) = E_\theta[Q/P_i]\) drain below one as atoms are added and
reweighted, while the right-hand panels trace the KL divergence, the gap, and
the guaranteed growth rate.

## Certifying the fit

The fitted mixture makes no validity claim on its own; `certify_trace()`
records the branch-and-bound search that bounds the worst-case null
expectation of `Q/P*` and thereby turns the fit into an e-variable.

```{r certify}
finished <- ripr_finish(
  state,
  reoptimise = TRUE, identify = TRUE, record_gap = TRUE
)
x <- likelihood(family(q), label = "Q") /
  likelihood(finished@P_star, label = "P*")
nodes <- certify_trace(x, plurality, tol = 1e-9)

vis_certify(nodes, alternative = family(q), part_labels = labels)
```

Each cell of the null is searched independently — the tabs choose a search,
the slider an iteration. The partition shades every live cell by its
certified upper bound, the panel beside it shows the enclosure
\([\mathrm{incumbent}, \mathrm{bound}]\) closing, and the scatter underneath
is the search tree, depth against birth iteration.

## Comparing fitting algorithms

`ripr` offers more than one way to add an atom. Frank–Wolfe takes the
maximiser of the field; Li–Barron scores each candidate by the KL of the
mixture *after* its weights are re-chosen, so each step achieves more and
costs several times as much. Which is ahead depends on whether steps or
seconds are being counted, and `vis_compare()` shows both: it overlays the
KL, gap and growth traces of several runs, with tabs switching the axis
between the step count and the clock. The clock is the trace's own
`elapsed` column -- the seconds each row cost -- accumulated.

```{r compare}
set.seed(1L)
state_lb <- ripr_init(family(q), plurality, record_gap = TRUE)
for (i in seq_len(6L)) {
  state_lb <- state_lb |>
    lb_step(record_gap = TRUE) |>
    em_step(record_gap = TRUE)
}

vis_compare(list(
  "Frank–Wolfe + EM" = state,
  "Li–Barron + EM" = state_lb
))
```

Only the trace is read, so the same comparison works for planar fits, for
`ripr_finish()` results, and for saved traces with no `ripr` installed.

## A Gaussian family with polyhedral nulls

The same verbs draw two-dimensional families whose
parameter is a point of the plane. Here the family is a bivariate Gaussian
with identity covariance, the alternative sits at the origin, and the null is
four pointy polyhedra -- each with four vertices and a single recession ray
pointing along an axis away from the origin, apex at distance one. The
viewport sizes itself to the parts' vertices (plus a margin), so the rays
simply run off the edge and the geometry near the alternative fills the view.

```{r gaussian}
gauss <- gaussian_family(d = 2L)
dart <- function(axis, sgn) {
  e <- c(0, 0)
  e[axis] <- sgn
  perp <- c(0, 0)
  perp[3L - axis] <- 1
  polyhedron_region(
    vertices = rbind(e, 1.75 * e + 0.6 * perp, 1.75 * e - 0.6 * perp,
                     2.5 * e),
    rays = rbind(e)
  )
}
darts <- null_model(
  gauss,
  list(dart(1, 1), dart(1, -1), dart(2, 1), dart(2, -1))
)
darts_labels <- c("+x", "-x", "+y", "-y")

vis_problem(darts, alternative = gauss(c(0, 0)), part_labels = darts_labels)
```

The sample space is continuous, so the fit runs on a quadrature engine, and
the field ships as one grid of values per step -- evaluated in R with the
same importance-weighted quadrature the fit itself used, so the gap markers
agree with the drawn field by construction. Starting the fit from each
part's far tail vertex (rather than the default projection of the
alternative) leaves the optimiser visible work to do.

```{r gaussian-fit}
set.seed(2L)
state2d <- ripr_init(
  gauss(c(0, 0)), darts,
  engine = gh_engine(24L),
  atoms = list(
    rbind(c(2.5, 0)), rbind(c(-2.5, 0)), rbind(c(0, 2.5)), rbind(c(0, -2.5))
  ),
  control = ripr_control(snapshot = "all")
)
for (i in seq_len(10L)) {
  state2d <- state2d |>
    fw_step(record_gap = TRUE) |>
    em_step(record_gap = TRUE)
}

vis_fit(state2d, part_labels = darts_labels)
```

## Figures for print

Each `vis_*()` verb has a `plot_*()` twin that draws one frame with base
graphics, so the usual devices and layouts apply. `plot_fit()` and
`plot_certify()` take the step or iteration to draw, defaulting to the last.

```{r static, fig.width = 7, fig.height = 7, out.width = "100%"}
op <- par(mfrow = c(2, 2), mar = c(1, 1, 2, 1))
plot_problem(state)
title("The problem")
plot_fit(state, step = 1)
title("The fit, at the start")
plot_fit(state)
title("The fit, at the end")
plot_certify(nodes, iteration = 3)
title("Certification, iteration 3")
par(op)
```

```{r static-compare, fig.width = 7, fig.height = 5, out.width = "100%"}
plot_compare(list(
  "Frank–Wolfe + EM" = state,
  "Li–Barron + EM" = state_lb
))
```

Wrap them in `pdf()` or `svg()` for a paper.

## Saving and redrawing

Every widget and plot draws from a plain-list payload. `riprvis_payload()`
extracts it, and any `vis_*()` or `plot_*()` verb of the same kind redraws
it, with or without `ripr` installed:

```{r payload, eval = FALSE}
saveRDS(riprvis_payload(vis_fit(state)), "fit.rds")

# Later, anywhere riprvis is installed:
p <- readRDS("fit.rds")
vis_fit(p)
plot_fit(p, step = 5)
```
